---
title: "Analyzing American Time Use Surveys in R"
subtitle: "An Example: Mothers' Time Use by Marital Status"
author: "Joanna R. Pepin"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output: html_document
---
# Getting the data
This tutorial uses data downloaded from IPUMS' [ATUS-X](https://www.atusdata.org/atus/) extract builder.
The data includes samples from 2003 - 2018.

## Variables:

Type  | Variable      | Label
----- | --------------| ----------------
H	    |	RECTYPE	      |	Record Type  
H	    |	YEAR	        |	Survey year  
H	    |	CASEID	      |	ATUS Case ID
H	    |	REGION        |	Region
H	    |	HH_SIZE	      |	Number of people in household
H	    |	HH_CHILD	    |	Children under 18 in household
H	    |	HH_NUMKIDS  	|	Number of children under 18 in household
H	    |	AGEYCHILD    	|	Age of youngest household child
H	    |	HH_NUMADULTS	|	Number of adults in household
H     | HHTENURE      | Living quarters owned, rented, or occupied without rent
P	    |	PERNUM	      |	Person number (general)
P	    |	LINENO	      |	Person line number
P	    |	DAY	          |	ATUS interview day of the week
P	    |	WT06        	|	Person weight, 2006 methodology
P	    |	AGE         	|	Age
P	    |	SEX	          |	Sex
P	    |	RACE         	|	Race
P	    |	HISPAN	      |	Hispanic origin
P	    |	MARST	        |	Marital status
P   	|	EDUC         	|	Highest level of school completed
P	    |	FULLPART     	|	Full time/part time employment status
P	    |	SPOUSEPRES	  |	Spouse or unmarried partner in household
P     | SPSEX         | Sex of respondent's spouse or unmarried partner
P	    |	HH_NUMOWNKIDS	|	Number of own children under 18 in household
P	    |	KIDUND13	    |	Own child under 13 in household
A	    |	ACTLINE	      |	Activity line number
A	    |	ACTIVITY	    |	Activity
A	    |	DURATION    	|	Duration of activity


Additionally, two time use variables were constructed in ATUSX:  

* The *telesolo* is a constructed time use variable from activity codes 120303 & 120304 conducted while *alone*.
* The *telesp* is a constructed time use variable from activity codes 120303 & 120304 conducted with a *spouse* or *unmarried partner* present.

Extract Notes:  

* DATA FORMAT: .dat (fixed-width text)
* STRUCTURE: Hierarchical
* SAMPLE MEMBERS: Respondents
  
## Data Download and Import
IPUMS NOTE: To load data, you must download both the extract's data and the DDI.  
Download the .DAT file after ATUSX processes your request (you will get an email).  
To download the DDI file, right click on the DDI link under the CODEBOOK heading, and select "save link file as"  

Detailed instructions for importing the data can be found at:   [https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html](https://cran.r-project.org/web/packages/ipumsr/vignettes/ipums.html).  
  
This R code was executed in the open source edition of [RStudio](https://www.rstudio.com/products/RStudio/).
  
### Set the working directory file path
You will need to change your file path to the local folder where you saved the ATUS data extracts.  
Notice the "/" lean to the right and not the left "\" if you are using a Windows computer.
``` {r eval = FALSE}
setwd("C:/Users/Joanna/Dropbox/MaritalStatus")
```

``` {r setup, echo = FALSE}
knitr::opts_knit$set(root.dir = 'C:/Users/Joanna/Dropbox/Sayer/MaritalStatus/Marital Status & Time Use --JP/Data Points')
```

### Load the libraries

```{r warning = FALSE, error = FALSE, message = FALSE}
library(ipumsr)
library(tidyverse)
```
Libraries can be installed with the commands `install.packages("ipumsr")` and `install.packages("tidyverse")`. These only need to be installed once but users must load the libraried each time Rstudio is started.

### Load ATUS Data into R
Users should change the `"atus_00040.xml"` to the file name of their own ATUSX extract.
``` {r message = FALSE}
ddi <- read_ipums_ddi("atus_00040.xml")
data <- read_ipums_micro(ddi)
```

# Set-up the Data

## Overview of the data
```{r eval = FALSE}
# Make sure data is now a dataframe.
class(data)

#Look at the structure of the data
summary(data)

# Check-out the variable names
names(data)
```

```{r results = 'hide'}
# Make the variable names lowercase
data <- data %>% rename_all(tolower)
```

### Change the variables' class from labelled to integer, character, or factor
``` {r results = 'hide'}
data <- data %>%
  mutate( year       = as.integer(lbl_clean(year)),
          caseid     = as.character(lbl_clean(caseid)),
          hh_size    = as.integer(lbl_clean(hh_size)),
          hh_child   = as.integer(lbl_clean(hh_child)),
          hh_numkids = as.integer(lbl_clean(hh_numkids)),
          ageychild  = as.integer(lbl_clean(ageychild)),
          hh_numadults = as.integer(lbl_clean(hh_numadults)),
          hhtenure   = as_factor(lbl_clean(hhtenure)),
          day        = as_factor(lbl_clean(day)),
          wt06       = as.integer(lbl_clean(wt06)),
          age        = as.integer(lbl_clean(age)),
          sex        = as_factor(lbl_clean(sex)),
          race       = as_factor(lbl_clean(race)),
          hispan     = as_factor(lbl_clean(hispan)),
          marst      = as_factor(lbl_clean(marst)),
          educ       = as.integer(lbl_clean(educ)),
          fullpart   = as.character(lbl_clean(fullpart)),
          spousepres = as_factor(lbl_clean(spousepres)),
          spsex      = as_factor(lbl_clean(spsex)),
          hh_numownkids = as.integer(lbl_clean(hh_numownkids)),
          kidund13   = as_factor(lbl_clean(kidund13)),
          activity   = as.character(lbl_clean(activity)),
          region     = as_factor(lbl_clean(region)),
          telesolo   = as.integer(lbl_clean(telesolo)),
          telesp     = as.integer(lbl_clean(telesp)))
```

### Prepare duration and activity variables
```{r results = 'hide'}
# Change NA to 0 for duration & activity minutes
data[["duration"]][is.na(data[["duration"]])] <- 0
summary(data$duration)

data[["activity"]][is.na(data[["activity"]])] <- "0"
summary(data$activity)
```

```{r}
# Check that duration = 1440
data %>%
  group_by(caseid) %>%
  summarise(total= sum(duration))
```

# Create activity variables by person

## Create the activity variable components
This creates a logical object in RStudio for each activity category.  
The activity codes can be found in ATUS'[Activity Coding Lexicon](https://www.bls.gov/tus/lexiconnoex0317.pdf).  

NOTE: The figure published in [PAA Affairs](http://www.populationassociation.org/wp-content/uploads/PAA-Spring19.pdf) used a more conservative 
measure of housework activities. This code includes all housework activities.  
```{r results = 'hide'}

# Childcare
ccare <- data$activity %in% 
  c(030100:030400, 080100:080200, 180381)

# Housework
hswrk <- data$activity %in% 
  c(020101:030000, 080200:080300, 080700:080800, 090100:100000, 160106, 070101, 180701, 180904, 180807, 180903, 080699, 160106)

# Leisure
leisure <- data$activity %in% 
  c(120100:130000, 130100:140000, 160101, 160102, 181200:181400)

# Sleep
sleep <- data$activity %in% 
  c(010100:020000)

#Leisure sub-types
socl <- data$activity %in% 
  c(120100:120200, 120200:120300, 120400:120500, 120501, 120502, 120504, 120599, 130200:130300, 130302, 130402)

actl   <- data$activity %in% 
  c(120307, 120309:120313, 130100:130200, 130301, 130401, 130499)

pass   <- data$activity %in% 
  c(120301:120306, 120308, 120399, 120503)

# Television
tele   <- data$activity %in% 
  c(120303, 120304)

```

## Create the activity identification variables in the dataset
Next, we use those logical objects to create the activity category variables in our dataset. First, we create a new variable called "actcat", with missing values. This variable will identify the activity categories, using the logical objects we created.
```{r results = 'hide'}
data$actcat<-NA
data$actcat[ccare]   <- "child care"
data$actcat[hswrk]   <- "housework"
data$actcat[leisure] <- "leisure"
data$actcat[sleep]   <- "sleep"
data$actcat <- as.character(data$actcat)
```

We'll also create a leisure category variable and one for television. Sub-categories of activties need their own variables. Here, television is a subcategory of passive leisure, which is a subcategory of leisure. Thus, we have more refined identification separated into different variables.
```{r results = 'hide'}
data$leiscat<-NA
data$leiscat[socl] <- "social leisure"
data$leiscat[actl] <- "active leisure"
data$leiscat[pass] <- "passive leisure"
data$leiscat <- as.character(data$leiscat)

data$telecat<-NA
data$telecat[tele] <- "television"
data$telecat <- as.character(data$telecat)
```
```{r echo = FALSE}
view <- select(data, caseid, activity, duration, actcat, leiscat, telecat)
knitr::kable(
  head(view, n = 20), 
  caption = "New Activity Category Variables.")
remove(view)
```

## Duration variables
Next we need to create summary variables with the total duration of each activty for each person
```{r results = 'hide'}
# Master activty variables
data <- data %>%
  group_by(caseid) %>%
  summarise (leisure  = sum(duration[actcat ==  "leisure"],    na.rm=TRUE),
             sleep    = sum(duration[actcat ==  "sleep"],      na.rm=TRUE),
             hswrk    = sum(duration[actcat ==  "housework"],  na.rm=TRUE),
             ccare    = sum(duration[actcat ==  "child care"], na.rm=TRUE)) %>%
  inner_join(data, by='caseid')

# Leisure activity variables
data <- data %>%
  group_by(caseid) %>%
  summarise (socl     = sum(duration[leiscat ==  "social leisure"],      na.rm=TRUE),
             actl     = sum(duration[leiscat ==  "active leisure"],      na.rm=TRUE),
             pass     = sum(duration[leiscat ==  "passive leisure"],     na.rm=TRUE)) %>%
  inner_join(data, by='caseid')

# Television variables
data <- data %>%
  group_by(caseid) %>%
  summarise (tele     = sum(duration[telecat ==  "television"],          na.rm=TRUE)) %>%
  inner_join(data, by='caseid')
```
```{r echo = FALSE}
view <- select(data, caseid, leisure, sleep, hswrk, ccare, socl, actl, pass, tele)
knitr::kable(
  head(view, n = 5), 
  caption = "New Activity Summary Variables.")
remove(view)
```

# Create tidy data
Currently, the data is in long format, with each person represented in multiple rows in the dataset. We will aggregate the data so each person is represented once per row.
```{r echo = FALSE}
view <- select(data, caseid, pernum, lineno, sex, actline, activity, sleep)
knitr::kable(
  head(view, n = 10), 
  caption = "Long format.")
remove(view)
```
  
Create a summary dataset for the variables that repeat each row, per person
```{r results = 'hide'}
max <- data %>%
  select(caseid, tele, socl, actl, pass, ccare, hswrk, leisure, sleep, year) %>%
  group_by(caseid) %>% 
  summarise_all(list(~max(.)))
```
Notice that each person is now represented only once per row.  
```{r echo = FALSE}
view <- select(max, caseid, tele, socl, actl, pass, ccare, hswrk, leisure, sleep)
head(view, n = 5)
remove(view)
```

Create a summary dataset for the variables that appear in the rectype = 1 rows, by person
```{r results = 'hide'}
rec1 <- data %>% 
  filter(rectype == 1) %>%
  select(caseid,  hh_size, hh_child, hh_numkids, ageychild, hh_numadults, hhtenure, region)
```
```{r echo = FALSE}
view <- select(rec1, caseid, hh_size, hh_child, hh_numkids, ageychild, hh_numadults, hhtenure, region)
head(view, n = 5)
remove(view)
```

Create a summary dataset for the variables that appear in the rectype = 2 rows, by person
```{r results = 'hide'}
rec2 <- data %>% 
  filter(rectype == 2) %>% 
  select(caseid, age, sex, race, hispan, marst, educ, fullpart, spousepres, spsex,  
         hh_numownkids, kidund13, day, wt06, telesolo, telesp)
```
```{r echo = FALSE}
view <- select(rec2, caseid, age, sex, race, hispan, marst, educ, fullpart, spousepres,  spsex,  
                     hh_numownkids, kidund13, day, wt06, telesolo, telesp)
head(view, n = 5)
remove(view)
```

Put them all together.
```{r results = 'hide'}
atus <- reduce(list(max, rec1, rec2), 
               left_join, by = "caseid")
```
Take a look at the person-level file with activity duration summary variables
```{r echo = FALSE}
head(atus, n = 5)
remove(max)
remove(rec1)
remove(rec2)
```

# Transform variables
```{r results = 'hide'}
# Gender
levels(atus$sex) <- c('Men', 'Women')

# Age
summary(atus$age)

# Marital status
  # cohabitors in one group regardless of marital history
  # use spousepres variable as indicator of marital status at time of ATUS diary
atus <- atus %>%
  mutate(
    mar = case_when(
      spousepres == "Spouse present"                                                         ~ "Married",
      spousepres == "Unmarried partner present"                                              ~ "Cohabiting",
      marst      == "Never married" & spousepres == "No spouse or unmarried partner present" ~ "Single",
      marst      != "Widowed" & marst != "Never married" & 
      spousepres == "No spouse or unmarried partner present"                                 ~ "Divorced/Separated", 
      TRUE                                                                                   ~  NA_character_ 
    ))
atus$mar <- factor(atus$mar, levels = c("Married", "Cohabiting", "Single", "Divorced/Separated"), ordered = TRUE)

# Spouse/partner sex
atus <- atus %>%
  mutate(
    spsex = case_when(
      spsex == "Male"                     ~ "Male",
      spsex == "Female"                   ~ "Female",
      spsex == "NIU (Not in universe)"    ~ "NIU",
      TRUE                                ~  NA_character_ 
    ))

atus$spsex <- factor(atus$spsex, levels = c("Male", "Female", "NIU"))

# Extended Family Member
atus <- atus %>%
  mutate(
    exfam = case_when(
      ((spousepres == "Spouse present" | spousepres == "Unmarried partner present") & hh_numadults <= 2) |
      ((spousepres == "No spouse or unmarried partner present") & hh_numadults <=1) ~ "No extra adults",
      ((spousepres == "Spouse present" | spousepres == "Unmarried partner present") & hh_numadults >= 3) |
      ((spousepres == "No spouse or unmarried partner present") & hh_numadults >=2) ~ "Extra adults",
      TRUE                        ~  NA_character_
    ))
atus$exfam <- factor(atus$exfam, levels = c("No extra adults", "Extra adults"), ordered = TRUE)

#Kid under 2
atus <- atus %>%
  mutate(
    kidu2 = case_when(
      ageychild <=2 ~ "Child < 2",
      TRUE          ~ "No children < 2"
    ))
atus$kidu2 <- as_factor(atus$kidu2)

# Number of own HH kids
summary(atus$hh_numownkids)

# Employment
atus <- atus %>%
  mutate(
    employ = case_when(
      fullpart == "1"  ~ "Full-time",
      fullpart == "2"  ~ "Part-time",
      fullpart == "99" ~ "Not employed",
      TRUE             ~  NA_character_
    ))
atus$employ <- factor(atus$employ, levels = c("Full-time", "Part-time", "Not employed"))

# Education
atus <- atus %>%
  mutate(
    educ = case_when(
      (educ >= 10 & educ <= 17)   ~ "Less than high school",
      (educ >= 20 & educ <= 21)   ~ "High school",
      (educ >= 30 & educ <= 31)   ~ "Some college",
      (educ >= 40 & educ <= 43)   ~ "BA or higher",
      TRUE                        ~  NA_character_
    ))
atus$educ <- factor(atus$educ, levels = c("Less than high school", "High school", "Some college", "BA or higher"))

# Race
atus <- atus %>%
  mutate(
    raceth = case_when(
      race   == "White only" & hispan == "Not Hispanic"  ~ "White",
      race   == "Black only" & hispan == "Not Hispanic"  ~ "Black",
      race   == "Asian only" & hispan == "Not Hispanic"  ~ "Asian",
      hispan != "Not Hispanic"                           ~ "Hispanic",
      TRUE                                               ~ "Other"
    ))
atus$raceth <- factor(atus$raceth, levels = c("White", "Black", "Hispanic", "Asian", "Other"))

# Weekend
atus <- atus %>%
  mutate(
    weekend = case_when(
      day == "Sunday"   | day == "Saturday"     ~ "Weekend",
      day == "Monday"   | day == "Tuesday" | day == "Wednesday" | 
        day == "Thursday" | day == "Friday"       ~ "Weekday",
      TRUE              ~  NA_character_
    ))
atus$weekend <- factor(atus$weekend, levels = c("Weekday", "Weekend"))

# Region
summary(atus$region)

# Home ownership
atus <- atus %>%
  mutate(
    ownrent = case_when(
      hhtenure == "Owned or being bought by a household member"     ~ "Own",
      hhtenure == "Rented for cash"                                 ~ "Rent",
      hhtenure == "Occupied without payment of cash rent"           ~ "Other",     
      hhtenure == "NIU (Not in universe)"                           ~ "Other",
      TRUE                                                          ~  NA_character_ 
    ))

atus$ownrent <- factor(atus$ownrent, levels = c("Own", "Rent", "Other"))
```

## Sample selection
Limit the dataset to White, Black, and Hispanic mothers of children with kids 13 or younger.  
Let's also restict the dataset to mothers who are prime working age.
```{r results = 'hide'}
atus <- filter(atus, sex == "Women") # women
atus <- filter(atus, spsex == "Male" | spsex == "NIU") # Different sex couples or singles
atus <- filter(atus, hh_numownkids >=1) # mothers
atus <- filter(atus, ageychild <=13) #mothers of kids 13 or younger
atus <- filter(atus, age >= 18 & age <=54) #prime working age
atus <- filter(atus, raceth != "Asian" & raceth != "Other") # white, black, and Hispanic
```

## Use listwise deletion to deal with missing data
```{r results = 'hide'}
# Keep only the variables of interest
atus <- select(atus, caseid, wt06, year, ccare, hswrk, leisure, sleep, socl, actl, pass, tele, telesolo, telesp, age, sex, mar, exfam, kidu2, hh_numownkids, employ, educ, raceth, weekend, region, ownrent)
# Listwise deletion
atus <- na.omit(atus)

# If want to look at only recent survey years, remove # below. Don't foget to then update the caption of the figure with the correct survey years.
# atus <- filter(atus, year >= 2012)
```
Here's a look at the data
```{r echo = FALSE}
str(atus)
```

# Create Linear Models and Margins
Load the `ggeffects` package in order to compute marginal effects from statistical models and save the results as tidy data frames.
```{r warning = FALSE, error = FALSE, message = FALSE}
library(ggeffects)
```
For each of activity variables of interest, run a linear model (`lm`) and then calculate the marginal effects (`ggeffect`)  
`ggeffect` calculates the adjusted predictions at the means, using proportions to fix factors

```{r results = 'hide'}
#Childcare
lm_care <- lm(ccare   ~ mar + exfam + hh_numownkids + kidu2 + educ + employ + raceth + age + weekend + region + ownrent, data = atus, weight=wt06)
pcare   <- ggeffect(lm_care, terms = "mar")

#Housework
lm_hswrk <- lm(hswrk  ~ mar + exfam + hh_numownkids + kidu2 + educ + employ + raceth + age + weekend + region + ownrent, data = atus, weight=wt06)
phswrk   <- ggeffect(lm_hswrk, terms = "mar")

#Leisure
lm_leis <- lm(leisure ~ mar + exfam + hh_numownkids + kidu2 + educ + employ + raceth + age + weekend + region + ownrent, data = atus, weight=wt06)
pleis   <- ggeffect(lm_leis, terms = "mar")

#Sleep
lm_sleep <- lm(sleep  ~ mar + exfam + hh_numownkids + kidu2 + educ + employ + raceth + age + weekend + region + ownrent, data = atus, weight=wt06)
psleep   <- ggeffect(lm_sleep, terms = "mar")
```

## Combine predictions into one dataframe
For each of the marginal effects datasets, label the group values as the activity
```{r results = 'hide'}
levels(pcare$group)[levels(pcare$group)=="1"]   <- "Childcare"
levels(phswrk$group)[levels(phswrk$group)=="1"] <- "Housework"
levels(pleis$group)[levels(pleis$group)=="1"]   <- "Leisure"
levels(psleep$group)[levels(psleep$group)=="1"] <- "Sleep"
```
Maniuplate the marital status variable
```{r results = 'hide'}
# Change the variable to be a factor variable from numeric
pcare$x <- as.factor(pcare$x)
phswrk$x <- as.factor(phswrk$x)
pleis$x <- as.factor(pleis$x)
psleep$x <- as.factor(psleep$x)
```
Combine the datatables
```{r results = 'hide'}
pred <- rbind(pcare, phswrk, pleis, psleep)
```
Prepare the categorical variables for data visualization
```{r results = 'hide'}
# Revalue the marital status factors to be readable
levels(pred$x)[levels(pred$x)=="3"] <- "Married"
levels(pred$x)[levels(pred$x)=="1"] <- "Cohabiting"
levels(pred$x)[levels(pred$x)=="4"] <- "Single"
levels(pred$x)[levels(pred$x)=="2"] <- "Divorced/Separated"

# Order the marital status and activity factors
pred$x     <- ordered(pred$x, levels = c("Married", "Cohabiting", "Single", "Divorced/Separated"))
pred$group <- ordered(pred$group, levels = c("Childcare", "Housework", "Leisure", "Sleep"))
```

Calculate the difference variable (mothers time use compared to married mothers time use, by activity)
```{r results = 'hide', warning = FALSE}
pred <- pred %>%
  group_by(group) %>%
  mutate(diff = predicted - predicted[x == "Married"])
```

# Data visualization
```{r fig.width=15,fig.height=10}
pred %>%
  filter(x != "Married") %>%
  ggplot(aes(x, diff, fill = x, label = round(diff, 0))) +
  geom_col() +
  facet_grid(~group) +
  ggtitle("Married Mothers Report More Housework and Less Leisure and Sleep Than Other Mothers") +
  labs(x = NULL, y = NULL, subtitle = "Predicted minutes per day with model controls",
       caption = "Source: American Time Use Surveys (2003 - 2018) \n Models control for extra adults, number of household kids, kids under 2,
       education, employment, race-ethnicity, age, weekend diary day, home ownership and region") +
  theme_minimal() +
  theme(plot.subtitle = element_text(size = 11, vjust = 1),
        plot.caption  = element_text(vjust = 1), 
        legend.position="none",
        strip.text.x  = element_text(size = 16),
        axis.title    = element_text(size = 14), 
        axis.text     = element_text(size = 12), 
        plot.title    = element_text(size = 16, face = "bold"),
        panel.grid.minor = element_blank(),
        panel.grid.major.x = element_blank()) +
  scale_y_continuous(breaks = c(-30, -20, -10, 0, 10, 20, 30), labels=c("-30", "-20", "-10", "Married mothers' \n minutes per day", "10", "20", "30")) +
  scale_fill_manual(values=c("#5D478B", "#CD3278", "#116A66")) +
  geom_errorbar(aes(ymin=diff-std.error, ymax=diff+std.error), width=.2,
                position=position_dodge(.9), color="grey")
```